---
title: "A Different Understanding of Democratic Constitutions?"
execute: 
  echo: false
---

```{r}
#| echo: false
#| label: wordstats
#| warning: false
#| message: false
library(wordcountaddin)
wordcount <- wordcountaddin::text_stats('Blog.qmd')
wordcount
```

## The tension between populism and liberal democracy

<!--# In their kick-off of this series, Petra Guasti and Mattia Zulianello briefly touched upon one of the most discussed features of populism: That it is in conflict with the core tenets of liberal democracy.   -->

Thinking of January 6th in the US, the PiS party's ignorance of judicial rulings in Poland, or Netanjahu's recent attempt to curb judicial power in Israel might easily make us regard populists as a danger to constitutional democracy. But, while these cases are extremely worrying, we should not infer from these salient cases to all populists in power. We are more likely to hear about all the cases in which populists overstepped constitutional boundaries than the many cases in which this is not the case.

In populism, the people's alleged homogenous general will is the only legitimate input in the political decision-making process. This of course clashes with the checks and balances, as well as minority rights that are inherent to liberal democracy and often established in a country's constitution. Often, this incompatibility is shortened to the claim that populism is bad for democracy and that populists in office will dismantle democratic institutions. But is that really true?

While this is often not mirrored in the public debate on populism, there has been a long-ongoing debate about whether populists share a different idea of what a constitution is supposed to be among scholars. Do populists understand it as a living document that changes with the people's interests and establishes their sovereignty, or do they simply use constitutional changes to consolidate power and dismantle democracy? Or is in the end a little of both? In our current research projects, we tested these claims empirically. And considering that leading scholars have emphasized populism's multiple facets for a long-time, our results are not surprising: It depends on the context!

## Varieties of Populism

Let's take a look at the bigger picture first. What happens around the world when populists change constitutions? To examine this, we look at the quality of liberal democratic institutions and participation in democracies in Europe and Latin America. With its people-centered approach, populism might have a positive effect on the inclusion of citizens in the democratic process while dismantling horizontal accountability. We use [V-Dem](https://v-dem.net/)'s participatory (PCI) and liberal component index (LCI) to operationalize democratic quality, and the [Comparative Constitutions Project](https://comparativeconstitutionsproject.org/)'s Database to code constitutional changes. To capture populism across two continents, we use [V-Party](https://www.v-dem.net/data/v-party-dataset/)'s populism score.

Our results show how different populists in power use constitutional changes to implement their understanding of democracy. While populists in power in Latin America have used constitutional changes to improve liberal democratic institutions - against our expectations -, they did not improve the quality of participatory democracy overall. In Europe, on the other hand, populists have used constitutional changes to decrease the quality of both democratic characteristics. You can check out different cases in the interactive tool and see what different effects constitutional changes by populists had on different indices of democratic quality. And since we all know that there is not one perfect way to code populism in parties, you can choose from different datasets.

```{r}
#| echo: false

vdem <- readRDS("data/ccpc_vdem_ela.rds")

ojs_define(data = vdem)

```

<!--# Why isn't Brazil coded as populist? -->

```{ojs}
//| echo: false

ojsdata = transpose(data)

viewof source = Inputs.select(
  ojsdata.map(row => row.country), 
  {label: "Select one", value: "Country", unique: true}
)

viewof populism = Inputs.select(
  ["Government Mean", "Government Weighted Mean", "Populist Prime Minister"], 
  {label: "How do you want do define populism?", value: "populism"}
)

filtered = ojsdata.filter(function(ojsdata) {
  return ojsdata.country == source;
})

constchange = filtered.filter(function(filtered) {
  return filtered.constchange > 0;
})

constchange2 = constchange.map(d => {
  let populism_score;
  if (populism === "Government Mean") {
    populism_score = d.gov_popul_mean;
  } else if (populism === "Government Weighted Mean") {
    populism_score = d.gov_popul_weighted;
  } else if (populism === "Populist Prime Minister") {
    populism_score = d.gov_popul_prime;
  } else {
    populism_score = null;
  }
  return {
    ...d,
    populism_score
  }
})


dashedLines = constchange2.filter(function(constchange2) {
  return constchange2.evnttype === 1;
})

solidLines = constchange2.filter(function(constchange2) {
  return constchange2.evnttype === 3;
})
```

::: panel-tabset
## Liberal Democracy

```{ojs}
console.log("dashedLines:", dashedLines);
console.log("solidLines:", solidLines);
console.log("constchange:", constchange2);

Plot.plot({
  marks: [
    Plot.ruleY([0]),
    ...constchange2.map(d => {
      return Plot.ruleX([d.year], {
        stroke: d.populism_score > 0.5 ? "#A44200" : "lightgrey",
        strokeDasharray: [5, 5]
      })
    }),
    ...solidLines.map(d => {
      return Plot.ruleX([d.year], {
        stroke: d.populism_score > 0.5 ? "#A44200" : "lightgrey",
        strokeDash: "solid"
      })
    }),
    Plot.lineY(filtered, {
      x: "year", 
      y: "v2x_liberal",
      // stroke = colr, if check in each row whether row.type = saved source then make red, otherwise lightgrey
      stroke: "black"      
    })
  ],
  y: {
    domain: [0, 1],
    label: "V-Dem Liberal Democracy Index"
  },
  x: {
    domain: [parseInt("1990"), parseInt("2021")],
    label: "Year",
    tickFormat: d3.format("d")
  }
})

```

## Participation

```{ojs, echo: false}

Plot.plot({
  marks: [
    Plot.ruleY([0]),
    ...constchange2.map(d => {
      return Plot.ruleX([d.year], {
        stroke: d.populism_score > 0.5 ? "#A44200" : "lightgrey",
        strokeDasharray: [5, 5]
      })
    }),
    ...solidLines.map(d => {
      return Plot.ruleX([d.year], {
        stroke: d.populism_score > 0.5 ? "#A44200" : "lightgrey",
        strokeDash: "solid"
      })
    }),
    Plot.lineY(filtered, {
      x: "year", 
      y: "v2x_cspart",
      // stroke = colr, if check in each row whether row.type = saved source then make red, otherwise lightgrey
      stroke: "black"      
    })
  ],
  y: {
    domain: [0, 1],
    label: "V-Dem Participation Index"
  },
  x: {
    domain: [parseInt("1990"), parseInt("2021")],
    label: "Year",
    tickFormat: d3.format("d")
  }
})
```
:::

<!-- To get a better sense of the changes of constitutional texts that happen under populist governments, we constructed indices for executive aggrandizement, judicial independence, political rights and social rights. Out of 89 constitutional changes under populist government, only 10 cases can be captured with these indices. -->

<!-- REFER TO RUTH-LOVELL ON IDEOLOGY! -->

If you go through the Latin American countries included in our study, you will see that the quality of participation did not decrease in every case. Still, our analysis shows that the general assumption that populism has a positive effect on participation does not hold. Neither does the widespread assumption that populists will harm liberal democratic institutions.

Our results underline the findings of many other papers that show how much variance there is in populist parties behavior toward democracy. To truly understand the effect of populism on democracy, we need to combine cross-country analyses with in-depth case studies. As a recent [paper](https://ejpr.onlinelibrary.wiley.com/doi/epdf/10.1111/1475-6765.12564) by Ruth-Lovell and Grahn shows, the effect of populism on democratic quality is conditioned by a country's previous democratic quality. It also depends on the party's host-ideology and of course its political strength.

To explain the varying effect of populism on constitutional democracy, we need to consider the circumstances in which the populist parties govern as well as the respective party's main ideology. In many multi-party systems populists are not in government by themselves, but govern with a mainstream coalition partner who forces them to [moderate](https://muse.jhu.edu/article/230454). In a [case study of Austria](https://ejpr.onlinelibrary.wiley.com/doi/epdf/10.1111/1475-6765.12573), we examine whether the right-wing populist FPÖ (and its splinter BZÖ) crossed constitutional boundaries more often than non-populists. They never hid their hostility towards the Austrian Constitutional Court in cases it ruled against them, calling it, for example, a "carnival joke". Earlier manifestos by the party also included a concept for a judicial reform that would have curbed the judiciary's power. Still, the party was not more likely to initiate legislation that was invalidated by the Austrian Constitutional Court. We argue that this is because their coalition partner pressured them into moderation the policy proposals, under the watchful eye of the public. When populists are in office, we should never forget to watch the ones they are governing with.

## Conclusion

In the end our research comes back to the conclusion that we need to consider the context populists govern in: What is their host-ideology? What majorities do they have?

Yes illiberalism is a challenge, but populism does not only consist of it and what facet of populism unfolds the most depends on its host-ideology.
