---
title: "Constitutional Regression or Different Understanding of Democratic Constitutions?"
execute: 
  echo: false
---

```{r}
#| echo: false
#| label: wordstats
#| warning: false
#| message: false
library(wordcountaddin)
wordcount <- wordcountaddin::text_stats('Blog.qmd')
wordcount
```

In their kick-off of this series, Petra Guasti and Mattia Zulianello touched upon one of the most discussed defining features of populism: That it is in conflict with two key features of liberal democracy: pluralism and minority rights. 

## Populists in power and constitutional regression - A myth?

The tension between populism and pluralism and minority rights is deeply rooted in the heart of populist ideology. The idea of a homogenous 'people' with a general will as the only legitimate input in the political decision-making process is at odds with the checks and balances and minority rights that are central to liberal democracy and often enshrined in a country's constitution. This often leads to the assumption that populists use constitutional changes to consolidate their power and dismantle democracy, a phenomenon often described as 'constitutional regression'.

Recent events such as January 6th in the US, the PiS party\'s disregard for judicial rulings in Poland, or Netanjahu\'s attempt to curb judicial power in Israel speak volumes to the threat populists can pose to constitutional democracy and yet there are voices that argue that the populist approach to constitutionalism is not undemocratic, but simply a different idea of democracy: one that views the constitution as a living document that changes with the people's general will. We argue that the relationship between populism and constitutional regression is anything but clear cut and requires further investigation.


## What the data tell us

As worrying as the events in the US, Hungary, Poland, or Israel with respect to quality of liberal democracy are, we should be careful not to infer a general causal relationship between populism and constitutional regression from these salient cases. We are more likely to hear about cases where populists have used (or tried to use) constitutional changes to dismantle liberal democracy than those where they have not.

What happens around the world when populists change constitutions? In our recent research, we analyzed the relationship between populist government, constitutional changes, and liberal democracy empirically. We focus on the quality of liberal democratic institutions and participation in democracies in Europe and Latin America. With its people-centered approach, populism might have a positive effect on the inclusion of citizens in the democratic process while dismantling horizontal accountability. We use [V-Dem](https://v-dem.net/)'s participatory (PCI) and liberal component index (LCI) to operationalize democratic quality, and the [Comparative Constitutions Project](https://comparativeconstitutionsproject.org/)'s Database to code constitutional changes. To capture populism across two continents, we use [V-Party](https://www.v-dem.net/data/v-party-dataset/)'s populism score.

Our results show how different populists in power use constitutional changes to implement their understanding of democracy. While populists in power in Latin America have used constitutional changes to improve liberal democratic institutions - against our expectations -, they did not improve the quality of participatory democracy overall. In Europe, on the other hand, populists have used constitutional changes to decrease the quality of both democratic characteristics. 

Below, we implemented an interactive tool that lets users choose between XXX different data sets on populism to explore the constitutional changes and their effect on different indices of democratic quality in variety of countries.

```{r}
#| echo: false

vdem <- readRDS("data/ccpc_vdem_ela.rds")

ojs_define(data = vdem)

```

<!--# Why isn't Brazil coded as populist? -->

```{ojs}
//| echo: false

ojsdata = transpose(data)

viewof source = Inputs.select(
  ojsdata.map(row => row.country), 
  {label: "Select one", value: "Country", unique: true}
)

viewof dataset = Inputs.select(
  ["V-Party", "PopuList", "Ruth-Lovell/Grahn"], 
  {label: "Based on which dataset would you like to define populism?", value: "dataset"}
)

viewof populism = Inputs.select(
  [], 
  {label: "How do you want do define populism?", value: "populism"}
)

```

```{ojs}

viewof dataset = Inputs.select(
  ["Datset 1", "Dataset 2", "Dataset 3"], 
  {label: "Which Dataset would you like to use?", value: "dataset"}
)

viewof variables = Inputs.select(
  [], 
  {label: "Which Variable would you like to inspect?", value: "variable"}
)

{const datasetOptions = 
  "Dataset 1": ["a", "b"],
  "Dataset 2": ["c", "d"],
  "Dataset 3": ["c", "f", "g"]
}

dataset.on("change", () => {
  variables.options = datasetOptions[dataset].options;
  variables.value = variables.options[0];
});
```

```{ojs}

// Define the available options for each dataset
{const datasetOptions = 
  "V-Party": ["Government Mean", "Government Weighted Mean"],
  "PopuList": ["Government Mean", "Populist Prime Minister"],
  "Ruth-Lovell-Grahn": ["Government Mean", "Government Weighted Mean", "Populist Prime Minister"]
}
```

```{ojs}

// Update the populism input options when the dataset input changes

dataset.on("change", () => {
  populism.options = datasetOptions[dataset].options;
  populism.value = populism.options[0];
});

filtered = ojsdata.filter(function(ojsdata) {
  return ojsdata.country == source;
})

constchange = filtered.filter(function(filtered) {
  return filtered.constchange > 0;
})

constchange2 = constchange.map(d => {
  let populism_score;
  if (populism === "Government Mean") {
    populism_score = d.gov_popul_mean;
  } else if (populism === "Government Weighted Mean") {
    populism_score = d.gov_popul_weighted;
  } else if (populism === "Populist Prime Minister") {
    populism_score = d.gov_popul_prime;
  } else {
    populism_score = null;
  }
  return {
    ...d,
    populism_score
  }
})


dashedLines = constchange2.filter(function(constchange2) {
  return constchange2.evnttype === 1;
})

solidLines = constchange2.filter(function(constchange2) {
  return constchange2.evnttype === 3;
})
```

::: panel-tabset
## Liberal Democracy

```{ojs}
console.log("dashedLines:", dashedLines);
console.log("solidLines:", solidLines);
console.log("constchange:", constchange2);

Plot.plot({
  marks: [
    Plot.ruleY([0]),
    ...constchange2.map(d => {
      return Plot.ruleX([d.year], {
        stroke: d.populism_score > 0.5 ? "#A44200" : "lightgrey",
        strokeDasharray: [5, 5]
      })
    }),
    ...solidLines.map(d => {
      return Plot.ruleX([d.year], {
        stroke: d.populism_score > 0.5 ? "#A44200" : "lightgrey",
        strokeDash: "solid"
      })
    }),
    Plot.lineY(filtered, {
      x: "year", 
      y: "v2x_liberal",
      // stroke = colr, if check in each row whether row.type = saved source then make red, otherwise lightgrey
      stroke: "black"      
    })
  ],
  y: {
    domain: [0, 1],
    label: "V-Dem Liberal Democracy Index"
  },
  x: {
    domain: [parseInt("1990"), parseInt("2021")],
    label: "Year",
    tickFormat: d3.format("d")
  }
})

```

## Participation

```{ojs, echo: false}

Plot.plot({
  marks: [
    Plot.ruleY([0]),
    ...constchange2.map(d => {
      return Plot.ruleX([d.year], {
        stroke: d.populism_score > 0.5 ? "#A44200" : "lightgrey",
        strokeDasharray: [5, 5]
      })
    }),
    ...solidLines.map(d => {
      return Plot.ruleX([d.year], {
        stroke: d.populism_score > 0.5 ? "#A44200" : "lightgrey",
        strokeDash: "solid"
      })
    }),
    Plot.lineY(filtered, {
      x: "year", 
      y: "v2x_cspart",
      // stroke = colr, if check in each row whether row.type = saved source then make red, otherwise lightgrey
      stroke: "black"      
    })
  ],
  y: {
    domain: [0, 1],
    label: "V-Dem Participation Index"
  },
  x: {
    domain: [parseInt("1990"), parseInt("2021")],
    label: "Year",
    tickFormat: d3.format("d")
  }
})
```
:::

<!-- To get a better sense of the changes of constitutional texts that happen under populist governments, we constructed indices for executive aggrandizement, judicial independence, political rights and social rights. Out of 89 constitutional changes under populist government, only 10 cases can be captured with these indices. -->

<!-- REFER TO RUTH-LOVELL ON IDEOLOGY! -->

If you go through the Latin American countries included in our study, you will see that the quality of participation did not decrease in every case. Still, our analysis shows that the claim that populism has a positive effect on participation does not hold. Neither does the widespread assumption that populists will harm liberal democratic institutions.

Our results speak to the findings of many other papers that show how much variance there is in populist parties behavior toward democracy. To understand the effect of populism on democracy, we need to combine cross-country analyses with in-depth case studies. As a recent [paper](https://ejpr.onlinelibrary.wiley.com/doi/epdf/10.1111/1475-6765.12564) by Ruth-Lovell and Grahn shows, the effect of populism on democratic quality is conditioned by a country's previous democratic quality. It also depends on the party's host-ideology and of course its political strength.

<!--
To explain the varying effect of populism on constitutional democracy, we need to consider the circumstances in which the populist parties govern as well as the respective party's main ideology. In many multi-party systems populists are not in government by themselves, but govern with a mainstream coalition partner who forces them to [moderate](https://muse.jhu.edu/article/230454). In a [case study of Austria](https://ejpr.onlinelibrary.wiley.com/doi/epdf/10.1111/1475-6765.12573), we examine whether the right-wing populist FPÖ (and its splinter BZÖ) crossed constitutional boundaries more often than non-populists. They never hid their hostility towards the Austrian Constitutional Court in cases it ruled against them, calling it, for example, a "carnival joke". Earlier manifestos by the party also included a concept for a judicial reform that would have curbed the judiciary's power. Still, the party was not more likely to initiate legislation that was invalidated by the Austrian Constitutional Court. We argue that this is because their coalition partner pressured them into moderation the policy proposals, under the watchful eye of the public. When populists are in office, we should never forget to watch the ones they are governing with. -->

## Conclusion

In the end our research comes back to the conclusion that we need to consider the context populists govern in: What is their host-ideology? What majorities do they have?

Yes illiberalism is a challenge, but populism does not only consist of it and what facet of populism unfolds the most depends on its host-ideology.
