---
title: "Constitutional Regression or Different Understanding of Democratic Constitutions?"
execute: 
  echo: false
---

```{r}
#| echo: false
#| label: wordstats
#| warning: false
#| message: false
# library(wordcountaddin)
# wordcount <- wordcountaddin::text_stats('Blog.qmd')
# wordcount
```

In their kick-off of this series, Petra Guasti and Mattia Zulianello touched upon one of the most discussed defining features of populism: That it conflicts with two key features of liberal democracy: pluralism and minority rights.

## Populists in power and constitutional regression - A myth?

The tension between populism and core liberal democratic values stems from a principle deeply ingrained in populist ideology: the concept of a homogeneous 'people' possessing a singular will as the exclusive legitimate input in the political decision-making process, contradicting the checks and balances inherent in liberal democracies. Aspects such as minority rights, pluralism, or power-sharing are central tenets of liberal democracy and are often enshrined in a country's constitution. Consequently, this leads to the presumption that populists use constitutional changes to consolidate their power and dismantle democracy, a phenomenon referred to as 'constitutional regression'.

Recent events, including January 6th in the US, the PiS party's disregard for judicial rulings in Poland, or Netanjahu's attempt to curb judicial power in Israel, exemplify the threat populists can pose to constitutional democracy. However, some voices contend that the populist approach to constitutionalism is not inherently undemocratic but rather presents an alternative conception of democracy - one that views the constitution as a living document that evolves in accordance with the people's general will. We argue that the relationship between populism and constitutional regression is far from straightforward and warrants further investigation.

## What the data tell us

While the events in the US, Hungary, Poland, and Israel undoubtedly raise alarms regarding the state of liberal democracy, we should avoid making broad generalizations about a direct causal relationship between populism and constitutional regression based on these salient cases for one important reason: We are more likely to to be exposed to instances where populists have used (or tried to use) constitutional changes to undermine liberal democracy than cases where such attempts have been non-existent.

What happens around the world when populists change constitutions? In a forthcoming paper, we analyzed the relationship between populist government, constitutional changes, and liberal democracy empirically. Our study examines the impact of constitutional modifications by populist and non-populist governments on the quality of liberal democratic institutions and citizen participation, focusing on Europe and Latin America. Given its people-centered approach, populism may positively influence the inclusion of citizens in the democratic process while simultaneously dismantling horizontal accountability. To operationalize democratic quality, we utilize [V-Dem](https://v-dem.net/)'s participatory (PCI) and liberal component index (LCI), while constitutional changes are coded using data from the [Comparative Constitutions Project](https://comparativeconstitutionsproject.org/)'s Database. To capture populism across two continents, we employ [V-Party](https://www.v-dem.net/data/v-party-dataset/)'s populism score.

Our shed light on how different populists in power use constitutional leverage constitutional amendments to manifest their respective visions of democracy. While populists in power in Latin America have used constitutional changes to improve liberal democratic institutions, they did not improve the quality of participatory democracy overall. In contrast, in Europe, populists in power employed constitutional changes that resulted in a decline in the quality of both democratic characteristics.

Below, we implemented an interactive tool that lets users choose between 3 different data sets on populism to explore the constitutional changes and their effect on different indices of democratic quality in various countries.

```{r}
#| echo: false
#| label: data-blog
#| message: false

library(tidyverse)

vdem <- readRDS("data/ccpc_vdem_eu_la.rds")

write.csv(vdem, "data/ccpc_vdem_populism.csv")

```

## Participation

```{=html}
<iframe width="100%" height="822.71875" frameborder="0"
  src="https://observablehq.com/embed/@jasminsworkspace/populism@1021?cells=viewof+source%2Cviewof+Democracy%2Cviewof+Dataset%2Cviewof+Vars%2Cprint%2Cplotfinal"></iframe>
```
<!-- To get a better sense of the changes of constitutional texts that happen under populist governments, we constructed indices for executive aggrandizement, judicial independence, political rights and social rights. Out of 89 constitutional changes under populist government, only 10 cases can be captured with these indices. -->

<!-- REFER TO RUTH-LOVELL ON IDEOLOGY! -->

Go through some of the different countries, and you will see a significant variance in how populists in office use constitutional changes, implying that it is not simply populist ideology that drives constitutional regression. Our results align with the conclusions of numerous other studies, emphasizing the substantial diversity in the behavior of populist parties toward democracy. Existing scholarship points to the [host-ideologies](https://www.cogitatiopress.com/politicsandgovernance/article/view/919), the political system or a [country's previous democratic quality](https://ejpr.onlinelibrary.wiley.com/doi/epdf/10.1111/1475-6765.12564) as explanations.

So far, much of our knowledge on the intersection between populism and constitutionalism comes from a few salient cases. To understand the effect of populism on democracy, scholars should combine large-n cross-country analyses with in-depth case studies, especially at the intersection between populism and legal studies, to understand how populism is implemented in the law. As a first step, we proposed to utilize [judicial review as a measure of populist government performance.](https://ejpr.onlinelibrary.wiley.com/doi/epdf/10.1111/1475-6765.12573) We anticipate and encourage additional research to delve into this vital intersection, as there is much more to uncover.

```{=html}
<!--
To explain the varying effect of populism on constitutional democracy, we need to consider the circumstances in which the populist parties govern as well as the respective party's main ideology. In many multi-party systems populists are not in government by themselves, but govern with a mainstream coalition partner who forces them to [moderate](https://muse.jhu.edu/article/230454). In a [case study of Austria](https://ejpr.onlinelibrary.wiley.com/doi/epdf/10.1111/1475-6765.12573), we examine whether the right-wing populist FPÖ (and its splinter BZÖ) crossed constitutional boundaries more often than non-populists. They never hid their hostility towards the Austrian Constitutional Court in cases it ruled against them, calling it, for example, a "carnival joke". Earlier manifestos by the party also included a concept for a judicial reform that would have curbed the judiciary's power. Still, the party was not more likely to initiate legislation that was invalidated by the Austrian Constitutional Court. We argue that this is because their coalition partner pressured them into moderation the policy proposals, under the watchful eye of the public. When populists are in office, we should never forget to watch the ones they are governing with. -->
```
